
\chapter{News}

The news this time is that there is an actual printed copy of the
\emph{QL Assembly Language} book available. This has been created,
with my blessings, by a QL Forum member named ``Tinyfpga'' who prefers
to read paper books as opposed to PDF files. I can't disagree!

The original, and always the latest PDF version (I keep fixing small
errors, grammar, spelling etc) will always be found at:

\href{https://github.com/NormanDunbar/QLAssemblyLanguageBook/releases/latest}{https://github.com/NormanDunbar/QLAssemblyLanguageBook/releases/latest}.

If you prefer the paper version, it's a ``print on demand'' thing.
``Tinyfpga'' has posted these instructions on the QL Forum:

\href{https://qlforum.co.uk/viewtopic.php?f=9&t=3294&start=40\#p35207}{https://qlforum.co.uk/viewtopic.php?f=9\&{}t=3294\&{}start=40\#{}p35207}

I've modified them slightly as it looks like the process has changed
a little since first being published.

\emph{As mentioned in an earlier post I decided to \textquotedbl{}publish\textquotedbl{}
(with permission) Norman Dunbar's PDF book titled 'QL Today's QL Assembly
Language Programming Series - Book one' as a real book.}

\emph{I decided to publish it as an A4 354 page, lay-flat book for
easy referencing. The book costs \textlira 7.04 plus \textlira 3.50
for postage and packaging direct from the printers. The method I have
used for very low cost one-off printing to order means that anyone
wanting to buy a book has to use my account to do so, as follows:}
\begin{itemize}
\item Go to \href{http://www.bookprintinguk.com}{www.bookprintinguk.com} 
\item Login as documentsforsms@gmail.com Password - forsms11 (That's two
digit ones on the end)
\item Click on the Image of the book
\item When the ``What would you like to do'' page comes up, click on the
``Order More'' link
\item Select \textquotedbl{}order one copy\textquotedbl{} and then go next
etc until you get to buy with your own name and address.
\end{itemize}
Hopefully, the process still remains the same.
