\chapter{Merry Christmas}

It's probably a little late to wish you a Happy Christmas but this year all my followers received a free download of a slightly updated eBook containing all the articles published in \emph{QL Today} over the past `n' years (longer than I care to think about!) 

All the diagrams etc have been converted from ASCII art to use proper png formats created with the ``graphviz'' utility (\url{http://www.graphviz.org/}). There was quite a bit of work involved in getting the old text, diagrams and code converted to \LaTeX (that's how they like it to be written!) but I think, so far, everyone thinks that a good job has been done. Even if I say so myself.

The eBook has been typeset using \LaTeX which is a professional system, much loved and used in academia and science for thesis\footnote{What is the plural of thesis?} and scientific papers. I even have a couple, probably more, technical manuals written and published using \LaTeX, my favourite being \emph{Compiler Design in C} by Allen Holub. Speaking of compiler design, there are some relevant bits coming up in this issue on this matter, but don't panic! 

If you have not already downloaded your free copy - it's in pdf format only at the moment, then please go to \url{http://qdosmsq.dunbar-it.co.uk/downloads/QLToday/QL_Assembly.pdf} and help yourself. I am looking into converting the pdf to other formats as some of my readers would like a version for Kindle and other eReaders.

Elsewhere in this issue, you will find some observations by George Gwilt on this eBook and its contents. George also has comments and observations on the first issue of this eMagazine, they are coming up next...

Happy new year, may all my readers enjoy a prosperous 2015.
