\chapter{Happy New Year}

Well then, 2020 was an absolute bummer\footnote{Yes, I \emph{am} being polite for once!} of a year. Hopefully we are all safe and well and have all survived, mentally and physically, through myriads of lockdowns and are following the rules to try and prevent another in a seemingly endless procession of lockdowns.

I'd like to wish you all a somewhat belated Happy New Year for 2021, let's hope it improves things a lot over 2020, although at the time of writing (January 13th) things aren't looking great. At least we have a vaccine or three.

It's been a while, I admit to my shame, since the last issue "hit the streets" back in September 2019, but I've been busy as some of you are aware. I had my first book published by Apress!

The book is called \emph{Arduino Software Internals}, and it covers how the Arduino Language works and how it connects with  the actual hardware of the Atmega328P/Atmega328AU microcontroller at the heart of the Arduino Board. It has taken me over two years to write but it's done now and out there in all good bookshops, on \href{https://www.amazon.co.uk/Arduino-Software-Internals-Complete-Language/dp/1484257898/ref=sr_1_1}{Amazon.co.uk} (other Amazons are available, just change the ``.co.uk'' to your chosen domain, and on \href{https://www.apress.com/gb/book/9781484257890}{Apress.com} too. (It's cheaper on Apress and you get free worldwide postage.)

Amazon had paper and Kindle versions, Apress has paper, PDF and EPUB versions.

The good news is, I'm writing another!

I hope you think that this issue of the ePeriodical was worth the wait.

