
\chapter{Feedback}

\section{GWASL }

This is feedback from myself!

GWASL is George Gwilt's 68008 assembler for the QL, I mentioned it
in Beginner's Corner last issue and since then, I've found that GWASL
will not assemble code written with the SMSQ/E trap codes and vector
names which they have a dot in their names, \texttt{OPW.SCR} for example.
George explicitly rejects names with dot's in for some reason, but
GWASS is fine with them.

I have the \emph{released} source code to GWASL, but unfortunately,
George has some library routines on his hard disc that I'm not privvy
to. So, I might have to do some disassembling to get a working source
that I might then be able to fix. George seems to have dropped off
the planet since not long after \emph{QL Today} folded.

I also found a couple of other bugs, which I might like to fix as
well:
\begin{itemize}
\item File names are limited to 40 characters, including the 5 character
device name, but I have a file with 42 characters (\texttt{win1\_source\_ReadingFiles\_ReadingFiles\_asm})
and I can't assemble it in situ, only when I copy it to ram1\_, for
example, and assemble it there.
\item Where a ``blank'' line actually consists of spaces, there's an ``illegal
instruction'' error which shouldn't be happening.
\end{itemize}
Anyway, because of this, Beginners' Corner now features the QMAC assembler
for those of us running on an original QL, or emulators, with a 68008
processor. GWASS will still be used for those with a 68020 processor,
although the code generated by Qmac will work fine on those too.
