\chapter{Using the MC68020 Instructions}
As you may be aware, in all of the articles I published in \textbf{QL Today} over the years, and in the preceding issues of this randomly occurring eMagazine, I've been a loyal user of George's \program{Gwasl}Gwasl assembler. This worked well on old black box QLs but who is using one of those these days? Anyone?

It is time to move on from the toys and playthings of childhood and become a real [wo]man. From the next issue, issue 4, we are going to switch to George's other assembler, \program{Gwass}Gwass and get down and dirty using the 68020 instructions. If you are using QPC, then you are already able to use them as George had a hand in getting QPC running on an emulated 68020 rather than a simple 68008 as the old Black Boxes used to run.

How many of my readers will this upset I wonder? \tablename{~\ref{tab:Emulators68020}} gives details of which computer or emulator can handle the new instructions.

\begin{table}[htbp]
\centering
\begin{tabular}{l l p{0.6\textwidth}}
\toprule
\textbf{Computer} & \textbf{Processor}&\textbf{Comments}  \\
\midrule
%
QL & 68008 & Cannot use the new instructions.\\
QPC & 68020 & Able to use the new instructions.\\
Others & 68008 & Cannot use the new instructions\\
%
\bottomrule
\end{tabular}
\caption{Emulators and the 68020}
\label{tab:Emulators68020}
\end{table}

This is a problem perhaps? Does anyone not use QPC for their main ``QL on a PC''? Would some or all of my readers be missing out if I went down this route?

You better let me know, soon(ish) at the usual email address \url{assembly@qdosmsq.dunbar-it.co.uk}.
